Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-03 paper/paper.tex
implemenatation

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

%\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}

@@

\section{Implementation}
% GC assumptions
In this section we describe in detail the process of liveness based garbage collection. 
During the execution of a program if it runs of out of memory, execution is stopped and the 
garbage collector is invoked. At this point all the references on the stack are checked for 
liveness and only live references are copied and the program continues execution. 
The main complexity in our algorithm is the way closures are handled. The following sub-sections 
show two diffrent ways of handling closures along with their pros and cons.
\subsection{A simple LGC Algorithm}
% Abstract description of the GC algorithm
\begin{verbatim}
procedure copy (cell)
 new_cell <-  dup_heapcell(cell)
 IF{celltype is cons}
     IF {cell.car is live}
        new_car        <-  copy(cell.car)
        new_cell.car   <-  new_car
     ENDIF
     IF {cell.cdr is live}
        new_cdr        <-  copy(cell.cdr)
        new_cell.cdr   <-  new_cdr
     ENDIF
 ELSE
  IF {celltype is closure}
    new_cell      <-  dup_heapcell(cell)
    IF {cell.arg1 is live}
       new_cell.arg1 <-  copy(cell.arg1)
    ENDIF
    IF {cell.arg2 is live}
       new_cell.arg2 <-  copy(cell.arg2)
    ENDIF
 ENDIF
 ENDIF
 copy_references_on_print_stack()
\end{verbatim}
In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
which might be created in a different environment(function). To overcome this we we carry some extra data 
with each closure, we carry the program point at which it was evaluated and also the names of the free 
variables contained in the closure. This gives us precise liveness information of the free variables of a 
closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
what is found at the creation point.

\subsection{Modified GC Algorithm}
Since the first algorithm requires a lot of bookkeeping and is not precise, we use a modified algorithm 
which does not use liveness for copying closures. 
\begin{verbatim}
procedure copy (cell)
 new_cell <-  dup_heapcell(cell)
 IF{celltype is cons}
     IF {cell.car is live}
        new_car        <-  copy(cell.car)
        new_cell.car   <-  new_car
     ENDIF
     IF {cell.cdr is live}
        new_cdr        <-  copy(cell.cdr)
        new_cell.cdr   <-  new_cdr
     ENDIF
 ELSE
  IF {celltype is closure && cell is not copied}
    new_cell      <-  dup_heapcell(cell)
    new_cell.arg1 <-  copy(cell.arg1)
    new_cell.arg2 <-  copy(cell.arg2)
  ENDIF
 ENDIF
 copy_references_on_print_stack()
\end{verbatim}

In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
assume everything under it to be live and copy everything that is reachable. Although this might 
lead to copying more cells it is faster as there are no liveness checks. Additionally we can 
optimize some of the processing by not traversing cells more than once. This is described in 
sub-section[]. \comment{Once experimental results are obtained compare the two algorithms here}
Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
and ensure references marked non-live are never copied using reachability. We discuss this issue 
and its solution in sub-section[].

\subsection{Optimization for RGC for closures}
One of the  reasosn why reachability based GC  is faster than liveness
based GC is  due to the fact that  in case we try to copy  a cons cell
which is  already copied  we do  not need to  traverse it  any further.
 The  same is not true in the case  of LGC, as the set
of  liveness paths  may not  be similar  in both  the  traversals. One
optimization is  to keep  track of all  the liveness  automaton states
that a particular  cons cell has been visited with  and ensure that we
do  not  traverse  it  with  the  same state  more  than  once.   This
optimization requires extra memory  for each heap cell and potentially
it can hold  all the automata states.  Therefore it  is not a feasible
solution. In case of lazy  languages since we use reachability to copy
closures, some cons  cells might be copied using  reachability. We can
avoid  repeated traversals of  these cons  cells without  tracking the
states. This only requires the  knowledge of whether the cons cell was
copied using  reachability.  Hence  we use a  flag to remember  if the
heap cell was  copied using reachability at any time, if  it is set we
do not traverse any further.\comment{Maybe add a diagram to make it clearer}

\subsection{Issues with mixing LGC and RGC}
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closures or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a
later GC
% and copy wrong locations. Avoiding it by setting non-live references
to null after every GC.
% Proof for the correctness.
\subsection{Garbage collection for references on print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's
book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied.
%% \begin{figure}
%% \begin{algorithm}[H]
%%  \KwData{this text}
%%  \KwResult{how to write algorithm with \LaTeX2e }
%%  initialization\;
%%  \While{not at end of this document}{
%%   read current\;
%%   \eIf{understand}{
%%    go to next section\;
%%    current section becomes this one\;
%%    }{
%%    go back to the beginning of current section\;
%%   }
%%  }
%%  \caption{How to write algorithms}
%% \end{algorithm}
%% \label{algo:LGC-Algo} \figrule
%% \end{figure}
%% We   provide   an   intuition    about   the   correctness   of   this
%% algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
%% about the print stack during a
%% GC,

>>>>

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

\hspace*{1cm}{\epsfig{file=motiv-example.eps, height=4cm}}

@@

\section{Implementation}
% GC assumptions
\cred{Introduction}

In this  section we describe in  detail the process  of liveness based
garbage collection.  During  the execution of a program  if it runs of
out  of memory,  execution is  stopped  and the  garbage collector  is
invoked. At this point all the references on the stack are checked for
liveness and only live references are copied and the program continues
execution.  The main  complexity in our algorithm is  the way closures
are  handled. The  following sub-sections  show two  different  ways of
handling closures along with their pros and cons.
\subsection{A simple LGC Algorithm}
\cred{Introduce terminology}
% Abstract description of the GC algorithm
\begin{verbatim}
procedure copy (cell)
 IF {cell is live}
   new_cell <-  dup_heapcell(cell)
   IF{celltype is cons}
        new_car        <-  copy(cell.car)
        new_cell.car   <-  new_car

        new_cdr        <-  copy(cell.cdr)
        new_cell.cdr   <-  new_cdr
    ENDIF
 ELSE
  IF {celltype is closure}
       new_cell.arg1 <-  copy(cell.arg1)
       new_cell.arg2 <-  copy(cell.arg2)
   ENDIF
 ENDIF
 copy_references_on_print_stack()
\end{verbatim}
In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
which might be created in a different environment(function), all of
which might not even exist on the program stack. To overcome this we we carry some extra data 
with each closure, we carry the program point at which it was evaluated and also the names of the free 
variables contained in the closure. This gives us precise liveness information of the free variables of a 
closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
what is found at the creation point. \cred{repeated traversals if sharing}.

\subsection{Modified GC Algorithm}
Since the first algorithm requires a lot of bookkeeping and is not
precise, we have a second version of the algorithm 
which does not use liveness for copying closures. 
\begin{verbatim}
procedure copy (cell)
 new_cell <-  dup_heapcell(cell)
 IF{celltype is cons}
     IF {cell.car is live}
        new_car        <-  copy(cell.car)
        new_cell.car   <-  new_car
     ENDIF
     IF {cell.cdr is live}
        new_cdr        <-  copy(cell.cdr)
        new_cell.cdr   <-  new_cdr
     ENDIF
 ELSE
  IF {celltype is closure && cell is not copied}
    new_cell      <-  dup_heapcell(cell)
    new_cell.arg1 <-  copy(cell.arg1)
    new_cell.arg2 <-  copy(cell.arg2)
  ENDIF
 ENDIF
 copy_references_on_print_stack()
\end{verbatim}

In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
assume everything under it to be live and copy everything that is reachable. Although this might 
lead to copying more cells it is faster as there are no liveness checks. Additionally we can 
optimize some of the processing by not traversing cells more than once. This is described in 
sub-section[]. \comment{Once experimental results are obtained compare the two algorithms here}
Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
and ensure references marked non-live are never copied using reachability. We discuss this issue 
and its solution in sub-section[]\comment{hanging like a sore thumb}.

\subsection{Optimization for RGC for closures}
One of the  reasosn why reachability based GC  is faster than liveness
based GC is  due to the fact that  in case we try to copy  a cons cell
which is  already copied  we do  not need to  traverse it  any further.
 The  same is not true in the case  of LGC, as the set
of  liveness paths  may not  be similar  in both  the
traversals. \cred{Push earlier: One
optimization is  to keep  track of all  the liveness  automaton states
that a particular  cons cell has been visited with  and ensure that we
do  not  traverse  it  with  the  same state  more  than  once.   This
optimization requires extra memory  for each heap cell and potentially
it can hold  all the automata states.  Therefore it  is not a feasible
solution.} In case of lazy  languages since we use reachability to copy
closures, some cons  cells might be copied using  reachability. We can
avoid  repeated traversals of  these cons  cells without  tracking the
states. This only requires the  knowledge of whether the cons cell was
copied using  reachability.  Hence  we use a  flag to remember  if the
heap cell was  copied using reachability at any time, if  it is set we
do not traverse any further.\comment{Maybe add a diagram to make it clearer}

\subsection{Issues with mixing LGC and RGC}
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closures or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a
later GC
% and copy wrong locations. Avoiding it by setting non-live references
to null after every GC.
% Proof for the correctness.
\subsection{Garbage collection for references on print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's
book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied.\cred{Very sketchy: Make
  another attempt to write}

%% \begin{figure}
%% \begin{algorithm}[H]
%%  \KwData{this text}
%%  \KwResult{how to write algorithm with \LaTeX2e }
%%  initialization\;
%%  \While{not at end of this document}{
%%   read current\;
%%   \eIf{understand}{
%%    go to next section\;
%%    current section becomes this one\;
%%    }{
%%    go back to the beginning of current section\;
%%   }
%%  }
%%  \caption{How to write algorithms}
%% \end{algorithm}
%% \label{algo:LGC-Algo} \figrule
%% \end{figure}
%% We   provide   an   intuition    about   the   correctness   of   this
%% algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
%% about the print stack during a
%% GC,
